Models

### 2021-05-31 analysis/05_model_metrics.Rmd
Restructuring

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(cowplot)
```

Get data and put it in long form 
```{r}
pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")

@@

#get metric data
metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  filter(!is.na(value)) %>%
  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  left_join(meta, c("rarefyID" = "rarefyid"))

@@

#find rarefyID's that have inf value for some metrics and exclude
cont_inf <- metrics %>% filter(is.infinite(value), !metric %in% c("qual.FRic",  "SES_qual.FRic")) %>% pull(rarefyID)

@@

model_data <- metrics %>% 
  filter(!rarefyID %in% cont_inf, !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup()

@@

####Fit GLM to overall metric trend

@@

Maybe need something other than a gaussian distribution?

@@

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("FRic", "FEve", "FDiv", "S")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = FALSE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

@@

@@

Summaries of study slopes
```{r}
#First - how many studys had a significant trend for each metric? 
study_slopes %>% 
  filter(sig == TRUE) %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
  group_by(metric) %>% 
  summarise(n_study_id = n_distinct(study_id)) %>%
  mutate(percent = n_study_id/54)

@@

```
Pairwise correlation analysis
```{r}
slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

@@

Plots of study-level slopes
```{r}
#pairwise plots using GGally
ggpairs(slopes_wide %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

@@

##Figures
Reproduce Dornelas figure - all data points & regressions w/overall mean
```{r}
#get climate classification
metadata <- read_csv(system.file("extdata", "biotime/biotime_metadata.csv", package = "biodivTS"))

fric_trend <- metrics %>%
  filter(metric == "FRic") %>%
  ggplot(aes(x = year, y = logvalue)) +
  # geom_smooth(aes(color = climate, group = plot_id), method = "glm", se = FALSE) +
  # geom_point(aes(color = climate, group = plot_id)) +
  geom_smooth(aes(group = rarefyID), color = "grey", method = "glm", se = FALSE) +
  geom_point(aes(group = rarefyID) ,color = "grey") +
  geom_smooth(color = "black", method = "glm", se = FALSE) +
  theme(legend.position = "none") 
ggsave("fric_trends.png", fric_plot)
```

```{r}
library(lme4)

plot_metric <- function(metric_name, ylabel, yval, data, model_coef){

>>>>

```{r}
library(tidyverse)
#library(moderndive)
library(lme4)
library(broom.mixed)
library(lmerTest)
library(cowplot)
```
#### Read in data, get final dataset for models
Get data and put it in long form 
```{r}
pins::board_register_github(repo = "karinorman/biodivTS_data", branch = "master")

@@

#get metric data
metrics <- pins::pin_get("rarefied-metrics", board = "github") %>%
  # pivot_longer(., cols = setdiff(everything(), one_of("rarefyID", "STUDY_ID", "SamplePool", "SampleN", "num_years", "duration", "startYear", "endYear", "YEAR", "cell", "rarefied", "type", "realm", "climate", "habitat", "biome_map", "taxa", "organisms", "cent_lat", "cent_long", "abundance_type", "BROAD_TYPE", "taxa_mod", "climate_mod")), names_to = "metric", values_to = "value") %>%
  rename_with(tolower, .cols = setdiff(everything(), one_of("rarefyID")))  %>%
  mutate(logvalue = log(value + 1)) %>% # create a column for log transformation that adds the absolute value integer closest to the most negative value to create all positive values
  left_join(meta, c("rarefyID" = "rarefyid"))

@@

@@

model_data <- metrics %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next",
                       "FRic", "FEve", "FDiv", "S"),
         !rarefyID %in% missing_null) %>%
  group_by(metric) %>%
  mutate(study_id =  str_extract(rarefyID, "[^_]+")) %>%
  mutate(year_scaled = scale(year), 
         scale_center = attributes(year_scaled)$`scaled:center`, 
         scale = attributes(year_scaled)$`scaled:scale`) %>%
  ungroup() %>%
  #this study has many low species richness observations, so FD metrics weren't calculated, just drop it
  filter(study_id != 348)

@@

####Fit GLMs to overall patterns and data broken down by groups

@@

@@

SES_rand_slope <- model_data %>%
  #filter(protected_area == "FALSE") %>%
  #filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "SES_FDis", )) %>%
  filter(metric %in%  c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "Jaccard_next")) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(value ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x)),
           # null_model = map(data, ~lmer(value ~ (year_scaled|study_id/rarefyID), 
           #                       data = .x)),
           # test = map2(model, null_model, anova),
           # test_res = map(test, tidy),
           coef = map(model, tidy),
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

#check that they all used the same number of observations
map(SES_rand_slope$model, nobs)

metric_rand_slope <- model_data %>%
  #filter(metric %in%  c("FRic", "FEve", "FDiv", "FDis", "S")) %>%
  filter(metric %in%  c("S")) %>%
  #it wants its own scaling to converge for some reason
  mutate(year_scaled = scale(year)) %>%
  group_nest(metric) %>%
  mutate(model = map(data, ~lmer(logvalue ~ year_scaled + (year_scaled|study_id/rarefyID), 
                                 data = .x, REML = TRUE)),
           coef = map(model, tidy)
           ) %>% 
  select(metric, model, coef) %>%
  unnest(cols = c(coef)) %>%
  filter(term == "year_scaled") %>%
  select(-c(effect, group, term, statistic))

#check that they all used the same number of observations
map(metric_rand_slope$model, nobs)

@@

#### Process Model Output

@@

Summaries of study slopes
```{r}
#First - how many studies had a significant trend for each metric? 
study_slopes %>% 
  filter(sig == TRUE) %>% 
  filter(metric %in% c("SES_FRic", "SES_FEve", "SES_FDiv", "Jaccard_base", "S")) %>% 
  group_by(metric) %>% 
  summarise(n_study_id = n_distinct(study_id)) %>%
  mutate(percent = n_study_id/54)

@@

#### Analizing slopes
Pairwise correlation analysis
```{r}
slope_corr <- corr.test(slopes_wide %>% ungroup() %>% select(starts_with("SES"), "S", "Jaccard_base"))

@@

@@

####Figures

Pairwise plots of study-level slopes
```{r}
#pairwise plots using GGally
ggpairs(slopes_wide %>% 
          select(-c("study_id", "FDiv", "FEve", "FRic")))

```{r}
plot_metric <- function(metric_name, ylabel, yval, data, model_coef){
